\bf Approachable case studies support learning and reproducibility in data science: An example from evolutionary biology

### 2022-06-10 jsdse/Manuscript.tex
fixing article and book title style on bib file

\bigskip
\begin{abstract}
Research reproducibility is essential for scientific development. Yet, rates of reproducibility are low, especially in the natural sciences. 
As increasingly more research relies on computers and software, efforts for improving reproducibility rates have focused on making available research workflows as computer code, as well as raw and processed data in computer readable form. 
However, research products that are digitally available are not necessarily friendly for learners and interested parties with little to no experience in the field. 
This renders research products unapproachable, counteracts availability, and hinders reproducibility short and long term. 
To improve long term adoption of reproducible workflows in research, they need to be made approachable for learners, the researchers of the future.

Using a case study within evolutionary biology, we identify aspects of research workflows that make them unapproachable to the general audience: 
use of highly specialized language; 
unclear goals and high cognitive load; and 
content-focused descriptions. 
We propose a set of principles to improve the unapproachable aspects of research workflows, and illustrate their application in a tutorial. 
We elaborate on the general application of these principles for documenting research workflows and products, to provide present learners and future researchers with tools for successful scientific reproducibility. 
%> ** add a sentence on implications for teaching reproducibility

%% Projects that address availability of code and data in the natural sciences
%% have focused on producing documentation that describes the nature and usage of the
%% resources.
%% However, it often communicates in a highly specialized language that is largely
%% inaccessible for the average target user. Additionally, baseline
%% computational knowledge and required skills to access scientific results are often
%% not found among potential users, and keep on increasing.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig-tree.png}
\end{center}
\caption{A phylogenetic tree from our tutorial. It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}. \label{fig:tree}}
\end{figure}

@@

The main goal of our case study is to obtain a single phylogeny summarizing data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.), our organism of study.
All analysis for our case study can be completely accomplished using functions from the R package \texttt{rotl} or the Python module \texttt{opentree}. 
If a researcher were to use the proposed analysis workflow in a publication, they would typically describe it in the methodology section as ``The canid summary phylogeny was obtained using functions from X package, details are available as supplementary files".
This is usual practice, mainly because journals do not have space to publish all code used for an analysis in the methods section. Yet, supplementary files and data have the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript \citep{pop2015use}.
They are also prone to the dreaded promise ``available upon request", which has very low rates of fulfillment \citep{krawczyk2012available}.
Without the primary data and code that was used to perform an analysis it is impossible to reproduce said analysis \citep{miyakawa2020no}. When the code is available, other issues can complicate reproduction of the analysis, to the point of completely obstructing reproducibility.
For example, some questions that are often unaddressed in research workflows are: what software do I need to open the code files?
Can I run the code with the same software I used to open the files, or do I need a different one?
Do I need additional software that the analysis depends on?
What does the code even mean?

These questions can usually be answered by referring to the software documentation, which are usually publicly available and can be accessed by any potential user.
As opposed to code, software documentation is written in natural language (i.e., any known human language, e.g., English, Spanish, Chinese), and is considered a key element for successful adoption of software by target users \citep{karimzadeh2018top}. This might explain why documentation for software addressed to academic users is also usually written using highly specialized computational language or jargon (i.e., computationally specific concepts, words, and phrases) as well as formal scientific and academic language.
We identify this as \textit{barrier 1} to approachable research workflows.-- \textit{Specialized language is intimidating}.
While scientific jargon might have an important role for formal acceptance of software by the scientific and academic community, it can be perceived as cold and/or intimidating language, that often slows down or even obstructs examination, application, and adoption of code by a wider audience \citep{ball2017its}.
In contrast, introducing information without the use of jargon, supports learners' conceptual understanding of new ideas and concepts \citep{mcdonnell2016concepts, pan2019online}.
% jargon creates a hostile environment that does not foster learning \citep{CITATION}.

@@

A third important aspect of software documentation are examples that demonstrate usage of individual functions \citep{karimzadeh2018top}.
Examples presented in software documentation are usually worked to perfection, as they are intended to showcase the ideal or minimal case in which a function works well.
Perfectly worked examples ignore the user experience by maintaining focus on the software content, and fail to provide users with expert and clear advice on how to troubleshoot if needed.
We identify this as \textit{barrier 3.-- Content-focused examples}.
Error management training is an approach that focuses on framing mistakes as beneficial to learning complex tasks, to give learners the opportunity to
actively explore a task with a positive mindset \citep{frese1995error}.
Providing examples that showcase potential errors supports users performance \citep{steele2014error}, and can greatly improve learners ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig-error.png}
\end{center}
\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:error}}
\end{figure}

@@

We propose that documents developed with ``literate programming'' can be made more accessible by choosing narratives that are relatable to a more general audience.
An easy way to do this in biology is choosing a charismatic taxon as a model organism.
For a research group this can be the group they are studying. For the general audience, a highly charismatic group such as dinosaurs should work well.
For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb} in January 2020, we tailored all examples to frogs and their allies.
% \citep{cite-aw-workshop}

@@

An activity that has become increasingly widespread in programming-language education is live programming. During live programming, and instructor writes code and executes it in a way that is visible to all learners through a screen \citep{guzdial2013lure, selvaraj2021live}.
One benefit of this practice is that typos and mistakes occur, normalizing them for learners. 
Watching an instructor handling errors, demonstrates learners on how to solve them when they are outside theclassroom \citep{shannon2015live, nederbragt2020ten}.
When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer with the Carpentries (a non-profit group that teaches foundational coding and data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
a senior faculty member taking the workshop complained that the typos were slowing things down and interfering with the pedagogy.
McTavish replied "the typos ARE the pedagogy".
This has become a slogan of sorts at the Carpentries, capturing the idea that embracing and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
Yet, working through mistakes is rarely done on written pedagogical materials.
Software documentation focuses on demonstrating usage function with examples that work seamlessly, without errors. 
We argue that the opposite is needed to support adoption of reproducible workflows and support long term independence in learners and users performance \citep{gaspar2007restoring, steele2014error}.

In our tutorials, we apply this principle by demonstrating examples that do not work as expected, and exemplifying ways to address them (Figure \ref{fig:error}).
For example, we identified inputs that would give a wide range of warnings and errors. We then focus on providing explanations for these errors and warning messages. We believe this supports users and learners to be less intimidated by the messages, and to practice taking useful information out of them.
% This is probably part of the computer science curriculum, but it is not at all
% part of the natural sciences curriculum.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig-schedule.png}
\end{center}
\caption{Snapshot of the home to our tutorial website, showing part of the schedule.
 Our tutorial website was constructed using The Software Carpentries workshop template \citep{swc-workshop-template}. \label{fig:schedule}}
\end{figure}

>>>>

\bigskip
\begin{abstract}
Research reproducibility is essential for scientific development. Yet, rates of reproducibility are low, especially in the natural sciences.
As increasingly more research relies on computers and software, efforts for improving reproducibility rates have focused on making research products digitally available, such as publishing analysis workflows as computer code, and raw and processed data in computer readable form.
However, research products that are digitally available are not necessarily friendly for learners and interested parties with little to no experience in the field.
This renders research products unapproachable, counteracts their availability, and hinders scientific reproducibility.
To improve both short and long term adoption of reproducible practices in research, research products need to be made approachable for learners, the researchers of the future.

Using a case study within evolutionary biology, we identify aspects of research workflows that make them unapproachable to the general audience:
use of highly specialized language;
unclear goals and high cognitive load; and
content-focused descriptions.
We propose a set of principles to improve the unapproachable aspects of research workflows, and illustrate their application using an online teaching resource.
We elaborate on the general application of these principles for documenting not only research products but also teaching materials, to provide present learners and future researchers with tools for successful scientific reproducibility.
%> ** add a sentence on implications for teaching reproducibility

@@

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig_tree.png}
\end{center}
\caption{A phylogenetic tree from our tutorial. It was extracted using OpenTree of Life resources \citep{opentreeoflife2019synth} wrapped in the \texttt{rotl} R package \citep{michonneau2016rotl}. \label{fig:tree}}
\end{figure}

@@

The main goal of our case study is to obtain a single phylogeny summarizing data from a set of published phylogenies for the canids (the family of dogs, coyotes, wolves, etc.), our organism of study.
All analysis for our case study can be completely accomplished using functions from the R package \texttt{rotl} or the Python module \texttt{opentree}.
If a researcher were to use the proposed analysis workflow in a publication, they would typically describe it in the methodology section as ``The canid summary phylogeny was obtained using functions from X package, details are available as supplementary files".
This is usual practice, mainly because journals do not have space to publish all code used for an analysis in the methods section. Yet, supplementary files and data have the misfortune to not be peer-reviewed as thoroughly (or at all) as the main manuscript \citep{pop2015use}.
They are also prone to the dreaded promise ``available upon request", which has very low rates of fulfillment \citep{krawczyk2012available}.
Without the primary data and code that was used to perform an analysis it is impossible to reproduce said analysis \citep{miyakawa2020no}. When the code is available, other issues can complicate reproduction of the analysis, to the point of completely obstructing reproducibility.
For example, some questions that are often unaddressed in research workflows are:
\textit{what software do I need to open the code files?}
\textit{Can I run the code with the same software I used to open the files, or do I need a different one?}
\textit{Do I need additional software that the analysis depends on?}
\textit{What does the code even mean?}

These questions can usually be answered by referring to the software documentation, which are usually publicly available and can be accessed by any potential user.
As opposed to code, software documentation is written in natural language (i.e., any known human language, e.g., English, Spanish, Chinese), and is considered a key element for successful adoption of software by target users \citep{karimzadeh2018top}. This might explain why documentation for software addressed to academic users is also usually written using highly specialized computational language or jargon (i.e., computationally specific concepts, words, and phrases) as well as formal scientific and academic language.
We identify this as \textit{barrier 1} to approachable research workflows.-- \textit{Specialized language is intimidating}.
While scientific jargon might have an important role for formal acceptance of software by the scientific and academic community, it can be perceived as cold and/or intimidating language, that often slows down or even obstructs examination, application, and adoption of code by a wider audience \citep{ball2017its}.
In contrast, introducing information without the use of jargon, supports learner's conceptual understanding of new ideas and concepts \citep{mcdonnell2016concepts, pan2019online}.
% jargon creates a hostile environment that does not foster learning \citep{CITATION}.

@@

A third important aspect of software documentation are examples that demonstrate usage of individual functions \citep{karimzadeh2018top}.
Examples presented in software documentation are usually worked to perfection, as they are intended to showcase the ideal or minimal case in which a function works well.
Perfectly worked examples ignore the user experience by maintaining focus on the software content, and fail to provide users with expert and clear advice on how to troubleshoot if needed.
We identify this as \textit{barrier 3.-- Content-focused examples}.
Error management training is an approach that focuses on framing mistakes as beneficial to learning complex tasks, to give learners the opportunity to
actively explore a task with a positive mindset \citep{frese1995error}.
Providing examples that showcase potential errors supports user's performance \citep{steele2014error}, and can greatly improve learner's ability to troubleshoot outside the classroom \citep{shannon2015live, nederbragt2020ten}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig_error.png}
\end{center}
\caption{Snapshot of a section of the tutorial website, where we demonstrate a common error. \label{fig:error}}
\end{figure}

@@

We propose that documents developed with ``literate programming'' can be made more accessible by choosing narratives that are relatable to a more general audience.
An easy way to do this in biology is choosing a charismatic taxon as a model organism.
For a research group, this can be the biological group they are studying. For the general audience, a highly charismatic group --such as dinosaurs, should work well.
For example, when we presented our tutorial to the Amphibia Web Organization \citep{van2002amphibiaweb} in January 2020, we tailored all examples to frogs and their allies.
% \citep{cite-aw-workshop}

@@

An activity that has become increasingly widespread in programming-language education is live programming.
During live programming, an instructor writes code and executes it in a way that is visible to learners through a screen \citep{guzdial2013lure, selvaraj2021live}.
One benefit of this practice is that typos and mistakes occur, normalizing them for learners.
Watching an instructor handling errors, demonstrates learners on how to solve them when they are outside the classroom \citep{shannon2015live, nederbragt2020ten}.
When co-author McTavish was a postdoc, teaching an introductory programming workshop as a volunteer with the Carpentries (a non-profit group that teaches foundational coding and data science skills to researchers worldwide) \citep{wilson2006swc, SWCwebsite},
a senior faculty member taking the workshop complained that the typos were slowing things down and interfering with the pedagogy.
McTavish replied "the typos ARE the pedagogy".
This has become a slogan of sorts at the Carpentries, capturing the idea that embracing and discussing mistakes is essential to teaching programming \citep{wilson2019teaching}.
Yet, working through mistakes is rarely done on written pedagogical materials.
Software documentation focuses on demonstrating usage function with examples that work seamlessly, without errors.
We argue that the opposite is needed to support adoption of reproducible workflows and support long term independence in learner's and user's performance \citep{gaspar2007restoring, steele2014error}.

In our tutorials, we apply this principle by demonstrating examples that do not work as expected, and exemplifying ways to address issues (Figure \ref{fig:error}).
For example, we identified inputs that would give a wide range of warnings and errors.
We then focus on providing explanations for these errors and warning messages.
We believe this supports users and learners to be less intimidated by the messages, and to practice taking useful information out of them.
% This is probably part of the computer science curriculum, but it is not at all
% part of the natural sciences curriculum.

@@

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig_schedule.png}
\end{center}
\caption{Snapshot of the home to our tutorial website, showing part of the schedule.
 Our tutorial website was constructed using the software workshop template from the Carpentries \citep{swc-workshop-template}. \label{fig:schedule}}
\end{figure}
